---
title: "Data visualization using the ggplot2 package"
format:
  revealjs:
    chalkboard: true
execute:
  echo: true
editor: visual
---

## Basic plotting

```{r}
data(ToothGrowth)
str(ToothGrowth)
```

## Scatterplots

```{r}
plot(ToothGrowth$len)
```

## Scatterplots

```{r}
plot(ToothGrowth$dose, ToothGrowth$len)
```

## Scatterplots

```{r}
plot(ToothGrowth$dose, ToothGrowth$len, 
     xlab = "Dose", ylab = "Tooth Length")
```

## Scatterplots

```{r}
plot(ToothGrowth$dose, ToothGrowth$len, 
     xlab = "Dose", ylab = "Tooth Length", 
     main = "Guinea Pig Tooth Growth by Dosage")
```

## Scatterplots

```{r}
plot(ToothGrowth$dose, ToothGrowth$len, 
     xlab = "Dose", ylab = "Tooth Length", 
     main = "Guinea Pig Tooth Growth by Dosage", 
     col = ToothGrowth$supp, pch = as.numeric(ToothGrowth$supp))
```

## Scatterplots

```{r}
plot(ToothGrowth$dose, ToothGrowth$len, 
     xlab = "Dose", ylab = "Tooth Length", 
     main = "Guinea Pig Tooth Growth by Dosage", 
     col = ToothGrowth$supp, pch = as.numeric(ToothGrowth$supp))

legend(1.5, 15, c("OJ", "VC"), col = 1:2, pch = 1:2)
```

## Scatterplots

```{r}
plot(ToothGrowth$dose, ToothGrowth$len, 
     xlab = "Dose", ylab = "Tooth Length",
     main = "Guinea Pig Tooth Growth by Dosage",
     col = ToothGrowth$supp, pch = as.numeric(ToothGrowth$supp), cex = 2)

legend(1.5, 15, c("OJ", "VC"), col = 1:2, 
       pch = as.numeric(ToothGrowth$supp), cex=2)
```

## Line Graphs

```{r}
plot(ToothGrowth$len, type = "l") # note type is the letter l for "line"
```

## Line Graphs

```{r}
plot(ToothGrowth$len, type = "b") # note type is the letter b for "both"
```

## Line Graphs

```{r}
plot(ToothGrowth$len, type = "l")
lines(ToothGrowth$len + 2)
```

## Line Graphs

```{r}
plot(ToothGrowth$len, type = "l")
lines(ToothGrowth$len + 2, lty = 3, lwd = 3, col = "darkgreen")
```

## Bar Charts

```{r}
plot(ToothGrowth$dose)
```

## Bar Charts

```{r}
table(ToothGrowth$dose)
```

## Bar Charts

```{r}
barplot(table(ToothGrowth$dose))
```

## Bar Charts

```{r}
props <- table(ToothGrowth$dose)/nrow(ToothGrowth)
props
```

## Bar Charts

```{r}
barplot(props)
```

## Bar Charts

```{r}
barplot(props, ylim = c(0, .4))
```

## Bar Charts

```{r}
barplot(props, ylim = c(0, .4), width = .2, col = c("blue", "white", "red"))
```

## Bar Charts

```{r}
barplot(props, xlim = c(0, .4),
  width = .2, col = c("blue", "white", "red"),
  horiz = TRUE)
```

## Histograms

```{r}
hist(ToothGrowth$len)
```

## Histograms

```{r}
hist(ToothGrowth$len, freq = FALSE)
```

## Histograms

```{r}
hist(ToothGrowth$len, freq = FALSE, breaks = seq(0, 35, 2))
```

## Histograms

```{r}
hist(ToothGrowth$len, freq = FALSE, breaks = 25)
```

## Histograms

```{r}
hist(ToothGrowth$len, freq = FALSE, breaks = seq(0, 35, 2),
  main = "Histogram of Tooth Growth", xlab = "Tooth Length",
  ylim = c(0, .07), col = "orange")
```

## Boxplots

```{r}
boxplot(ToothGrowth$len)
```

## Boxplots

```{r}
boxplot(ToothGrowth$len~ToothGrowth$supp)
```

## Boxplots

```{r}
boxplot(ToothGrowth$len~ToothGrowth$supp,
  col = c("darkgreen", "purple"),
  ylab = "Supplement Type", xlab = "Tooth Length",
  ylim = c(0, 40), horizontal = TRUE)
```

## Multiple plots in one window

```{r}
par(mfrow = c(1, 2))
boxplot(ToothGrowth$len~ToothGrowth$supp, main = "Boxplot",
xlab = "Supplement Type")
  
hist(ToothGrowth$len, main = "Histogram", xlab = "Tooth Length")
```

## Multiple plots in one window

To reset plotting device run:

`dev.off()`

or

`par(mfrow = c(1, 1))`

## Saving Plots

```{r, eval=F}
# note: run getwd() to see the working directory -
# that is the directory to which files will be written
pdf("plot1.pdf", width = 6, height = 9)
boxplot(ToothGrowth$len~ToothGrowth$supp,
  col = c("darkgreen", "purple"),
  xlab = "Supplement Type", ylab = "Tooth Length",
  ylim = c(0, 40), horizontal = TRUE)
dev.off()

pdf("plot2.pdf", width = 4, height = 8)
boxplot(ToothGrowth$len~ToothGrowth$supp,
  col = c("darkgreen", "purple"),
  xlab = "Supplement Type", ylab = "Tooth Length",
  ylim = c(0, 40), horizontal = TRUE)
dev.off()
```

## Saving Plots

```{r, eval=F}
png("plot1.png", units = "in", res = 300,
  width = 6, height = 9)

boxplot(ToothGrowth$len~ToothGrowth$supp,
  col = c("darkgreen", "purple"),
  xlab = "Supplement Type", ylab = "Tooth Length",
  ylim = c(0, 40), horizontal = TRUE)

dev.off()
```
